Modal Abstractions for Virtualizing Memory Addresses

### 2025-07-21 programlogic.tex
C pass through Ismail changes: adding change-tracking macros, fixing bad suggestions from grammar tool

We describe a program logic (a separation logic) along the lines suggested earlier, where every assertion is relative
to an address space in which it is interpreted, allowing us to define \emph{virtual points-to} assertions that make claims
about memory locations in a particular address space. Virtual addresses, and even virtual points-to assertions, 
are not tagged with their address spaces in any way. Memory access in this logic is validated through the use
of virtual points-to assertions in preconditions, which guarantee that address translations succeed.
This supports rules for updating not only typical data in memory that happens to be subject to address translation but also
manipulation of the page tables themselves via virtual addresses (as demanded by all modern hardware) and also
via virtual points-to assertions.
To support specifications that deal with multiple address spaces, our logic incorporates a hybrid-style modality
$[r](P)$ to state that an assertion is true in another address space (assertion-specified) rather than the address space
currently active in hardware, which is not only useful for virtual memory manager invariants, but also critical to reasoning
about change of address space.
By developing this within the \iris framework, we obtain additional features (e.g., fractional permissions) that allow us to verify
some of the most subtle and technically challenging instruction sequences in an OS kernel (Section \ref{sec:experiment}).

% We derive a program logic (a separation logic) supporting the following stances and constraints:
% \begin{enumerate}
% \item \textit{address-spaces as modal contexts}: Assertions in our logic are context-dependent,
%   in the sense that their truth depends on which address space they are used in, due to the need to support virtual points-to assertions.
% \item \textit{sharing}: The physical location backing a virtual address's storage is located (during a page table walk) through a 
%       set of physical page-table (L4-L1 page-tables) acceses that are shared amongst different virtual addresses (specifically,
%       those on the same page of memory\footnote{of in the case of L2 or higher levels of tables, within a given broader region.}).
%       This sharing imposes constraints on defining points-to assertions
%       in terms of physical (L4-L1) page-table memory accesses
% \item \textit{context-agnostic-resources}: each virtual address is valid under a certain address-space, 
%       but it does not represent this \textit{knowledge} of its address-space. 
%       That is, assertions are not explicitly tagged with their address space validity
% \item \textit{updating address-space mappings}: We present logical abstractions to enable 
%       updating not only pages of typical data in memory, but also page tables themselves.\footnote{Prior work relied on unfolding operational semantics
%       to verify page table updates.}
% \item \textit{explicitly-modal assertions}: Our logic includes a means to talk about facts being true
%       in another address space
% \item \textit{address-space switch as changing the "World" of truth}: Switching from one address-space to another logically
%       becomes a simultaneous introduction-and-elimination of a pair of modal assertions (for different address spaces)
% \end{enumerate}

% The idea that the truth of an assertion is relative to an address space has far-reaching consequences.
To support making assertions depend on a choice of address space, we work entirely in a pointwise lifting of \iris's base BI logic,
essentially working with separation logic assertions indexed by a choice of page table root as a $\mathcal{W}_{64}$, which we call $\textsf{vProp }\Sigma$:\footnote{
  \iris experts may notice our \lstinline|-b>| resembles another pointwise lifting already in  \iris~\cite{dang2019rustbelt,dang2022compass}. 
  This similarity is real, but the existing lifting does not appear to work with indexed \coq types like our \lstinline|word n| as a domain.
}
\lstinline[language=Coq]|Definition vProp  $\Sigma$ : bi := word 64 -b> iPropI  $\Sigma$|.
This is the (\rocq) type of assertions in our logic.
Most constructs in \iris's base logic are defined with respect to any BI-algebra (of \coq type \lstinline|bi|), so they automatically
carry over to our derived logic.
\add{Since \textsf{vProp}s are ``just'' \textsc{Iris} predicates on a choice of page table root,
it may not be obvious what the gain is over simply using \textsc{Iris} directly.
The gain is not expressive power.
Every modal logic can be translated straightforwardly to a logic without modalities
(nearly always a small variation on the \emph{standard translation} from propositional modal logic
to first-order logic~\cite{blackburn2006modal}\footnote{\add{Itself similar to verification condition generation
for weakest precondition calculi and Hoare logics.}}) that can directly address the logic's models ---
but these translations, in addition to making specifications far more verbose,
generally do not preserve the appealing reasoning principles of modal logics.
This becomes more apparent when discussing our proof rules.
}

@@

\paragraph{Register points-to} 
The assertion $\textsf{r}\;\mapsto_{\textsf{r}}\{q\}\;\textsf{rv}$ ensures the ownership of the register $\rg$ containing the value of the register $\rv$. 
The fraction $\qfrac$ with value 1 asserts the unique ownership of the register mapping and grants update permission to it;
otherwise, any value $0 < \qfrac <1$ represents partial ownership, granting read-only permission on the mapping.

@@

\lstinline|L4_L1_PointsTo| works by
chaining together the entries for each level, using the sequence of table offsets from the address being translated to index
each table level, and using the physical page address embedded in each entry. \footnote{
  The fractions \lstinline|q1| through \lstinline|q4| represent the fractional ownership of each entry based on how many
  word-aligned addresses might need to share the entry ---  $(\frac{1}{512})^n$ for each level $n$.
}
For example, the first-level address translation to get the L4 entry (\lstinline|l4e|) 
  uses the masks \textsf{l4M52} with the current \lstinline|cr3| to get the physical address of the start of the L4 table
  and \textsf{l4off} with the virtual address being translated to compute the correct offset within that table.
  Then, that physical location is the appropriate entry in the L4 table \textsf{l4e}.
  Subsequent levels of the page table walk work similarly.
The statement of these assertions is simplified by the use of our split physical points-to assertions, since
each level of tables is page-sized. \footnote{We do not address superpages and hugepages in this paper.}
This helper definition is also more explicit in \textsf{vProp} which binds a value to \lstinline|cr3| and uses it to start the translation process.
\looseness=-1

@@

  We separate the physical page-table walk from the virtual points-to relation, replacing it with a ghost state that merely guarantees that the address translation would succeed.
  \iris includes a \emph{ghost map} construction, which we use to track mappings from virtual addresses to the physical addresses they translate to as a piece of ghost state.
  The map includes, for each key in the map (i.e.,
  each virtual address), a token $\ghostmaptoken{\gamma}{k}{v}$ whose ownership is required to update that key-value pair in the ghost map named $\gamma$. The existence of such a token implies that the actual map $\theta$ tracked by a corresponding $\mathsf{GhostMap}(\gamma,\theta)$
  resource indeed maps $k$ to $v$. These properties are captured by key \iris rules in Figure \ref{fig:ghostmaps}.\footnote{\iris ghost maps lack established notation in the literature,
  but the syntax we use captures the details of \texttt{iris.base\_logic.lib.ghost\_map}.}
  There are other rules, but these two are most important for explaining ghost maps.
  \textsc{GhostMapUpdate} says that ownership of the actual ghost map with ghost name $\gamma$ and map contents $\theta$,
  and a full (nonfractional) token witnessing that $\theta$ maps \textsf{pa} to \textsf{va} permits an update to the ghost map's state,
  changing the map and replacing the token to represent the new value.
  \textsc{GhostMapLookup} allows using the same information to simply conclude that the mapping indicated by the token is true.

@@

  % In Iris this is realized by using an authoritative resource algebra: there is a single \emph{authoritative} global copy of the (ghost)
  % map caching virtual-to-physical address translations, and for each entry a read only \emph{partial} ownership of that key-value pair.
  % The resource algebra itself is instantiated as:
  % \[\mathcal{A}\textsf{bsPTableWalk} \stackrel{\triangle}{=} \textsc{Auth} (\; \mathcal{W}_{64} \;\rightarrow_{\textrm{fin}} \;  ( (\textsc{Frac }, \mathord{+}) \times (\textsc{Agree } \Loc,\mathord{=}) ))\]
For clarity, we refer to the specific ghost map summarizing virtual-to-physical translations by \\
\centerline{$\mathcal{A}\textsf{bsPTableWalk}(\delta,\theta) \stackrel{\triangle}{=} \mathsf{GhostMap}(\delta,\theta)$}\\
(omitting $\delta$ for brevity when only one is in scope)
and keep this in a per-address-space invariant described shortly.
We then replace the physical traversal $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in Figure \ref{fig:strongvirtualpointsto}'s virtual points-to definition
with ownership of the token \ghostmaptoken{\delta}{\vaddr}{\paddr}, %ghost-map ($ \sumwalkabs\vaddr\qfrac\paddr$),
yielding Figure \ref{fig:virtualpointstosharing}'s definition.
This new definition guarantees that the ghost map contains an mapping from the virtual address ($\vaddr$) to a physical address ($\paddr$),
and thus that the per-address-space invariant will contain the physical resources that guarantee that the hardware resolves the translation.

  \begin{figure*}
  \footnotesize
\centerline{$
\begin{array}{l}
  \mathcal{I}\textsf{ASpace}(\ptablestore,m)\stackrel{\triangle}{=} \textsf{ASpace\_Lookup}(\ptablestore,m) \ast \\
 \qquad\qquad\qquad \bigast{(\vaddr, \textsf{paddr})\in \ptablestore}{\exists\;(\textsf{l4e l3e l2e, l1e, paddr})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e, paddr})} \\
  \textsf{ where } 
   \textsf{ASpace\_Lookup} (\ptablestore,m) \stackrel{\triangle}{=} \lambda\textsf{ cr3val} \ldotp \; \exists \gammaPred \; \ldotp \ulcorner m \; !!\; \textsf{cr3val} = \textsf{Some } \gammaPred \urcorner \ast
    %\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}
    \ptableabswalk{\delta,\theta}
\end{array}
$}
\vspace{-1em}
\caption{Per-address-space invariant with a fixed global map of address-space names $m$}
  \label{fig:peraspaceinvariant}
\vspace{-1em}
  \end{figure*}

We place the authorative ownership of the ghost translation $\mathcal{A}\textsf{PTableWalk}$ in a per-address-space invariant
$\mathcal{I}$\textsf{ASpace} (Figure \ref{fig:peraspaceinvariant}). 
$\mathcal{I}$\textsf{ASpace} alone allows changes to the page tables
that preserve overall virtual-to-physical translations.
When combined with the
fragment stored in virtual points-to (Figure \ref{fig:virtualpointstosharing}),
changes in the virtual-to-physical translations become possible without the need to track all virtual points-to assertions.

\todo[inline]{This next paragraph below explains details, but should first explain the big picture: the current address space
is identified by a paddr, so to state the invariant for the address space named by that paddr we need to look up
what invariants should hold, then assert that those invariants do hold.}
In Figure \ref{fig:peraspaceinvariant},
$\theta$ is the logical map from virtual addresses to the physical addresses they should translate to, 
for the \emph{current} address space, corresponding to the page tables currently installed (the ones
indicated by \lstinline|cr3|). In Figure \ref{fig:virtualpointstosharing}, the existentially quantified $\delta$
is the ghost name for this map (again, for the current address space).
Because our logic is the first to address the existence of multiple address spaces,
it must ensure that $\theta$ actually corresponds to the intended ghost map for the current address space.
This requires looking up the ghost name (i.e. $\delta$) for the current address space,
which is handled by $m$ --- another logical map (tied to ghost state shortly)
from various page table roots to ghost names for the per-address-space ghost maps discussed above.
$\mathcal{I}\textsf{ASpace}(\theta,m)$ then stores the physical resources for the current address space mappings,
ensuring that the current address space is valid (has an entry in $m$ whose corresponding
authoritative map matches $\theta$), and asserting ownership of the appropriate fractional ownership
of the physical page table walk for each virtual-to-physical translation in $\Theta$ (for the current address space).

@@

\subsection{Address-Space Management}
\label{sec:aspacemanagement}
% So far, we have introduced logical abstractions for a single address space, but VMMs
Real VMMs must
 handle more than one address-space.
Doing so requires a way to talk about other address spaces, and means to switch address spaces.
\begin{figure}
\footnotesize
\begin{mathpar}
    \mbox{$[r](\mathsf{P}) \;: \mathsf{vProp} \; \Sigma \; \stackrel{\triangle}{=} \; \lambda \_,\; \mathsf{P}~\mathsf{r} \vspace{0.5em}$}
    \and
    \mbox{$\textsf{Fact P} \stackrel{\triangle}{=} \;\forall \; ,r \; r' \ldotp  \textsf{P r} \dashv\vdash \textsf{P r'}$}
    \and
    \inferrule*{ }{\textsf{Fact}\;[r](\textsf{P}) }
    \and
    \inferrule*{ }{\textsf{Fact}\;(\textsf{r}\;\mapsto_{\textsf{r}}\{q\}\;\textsf{rv})}
    \and
    \inferrule{ }{\textsf{Fact}\;(\textsf{w} \mapsto_{\textsf{p}} \{q\} \textsf{ v} )}
    \and
    \inferrule{ }{(\textsf{P} \vdash \textsf{Q}) \vdash  ([r](\textsf{P}) \vdash  [r](\textsf{Q}))}
    \and
    \inferrule{ }{[r](\textsf{P} \ast \textsf{Q}) \dashv\vdash ([r](\textsf{P}) \ast [r](\textsf{Q})}\vspace{0.5em}
    \and
    \inferrule{ }{[r](\textsf{P}\land \textsf{Q}) \dashv\vdash [r](\textsf{P})\land [r](\textsf{Q})}
    \and
    \inferrule{ }{[r](\textsf{P}\lor \textsf{Q}) \dashv\vdash [r](\textsf{P})\lor [r](\textsf{Q})}
    \and
    \inferrule{ }{\textsf{Fact P} \vdash  [r](\textsf{P}) \dashv\vdash (\textsf{P})} %(\textsf{vProp }\Sigma))\\
\end{mathpar}
\vspace{-1em}
  \caption{Other-space Modality and Its Laws}
  \label{fig:modaldef}
\vspace{-1em}
  \end{figure}
Figure \ref{fig:modaldef} gives the definition of our modal operator for asserting the truth of a modal
(address-space-contingent) assertion \emph{in another address space}, which we call
the \emph{other-space} modality. The definition itself is not
particularly surprising --- as our modal assertions are semantic predicates on a page table root (physical)
address, the assertion $[r](P)$ is a modal assertion that ignores the (implicit) current page table root,
and evaluates the truth of $P$ as if $r$ were the page table root. 
The novelty here is not in the details of the definition, but in recognizing that this is the right way to deal with
multiple address spaces, and working out how to support interaction of multiple address spaces (discussed in the next section).
\add{The modal assertions, together with the other-space modality, mean we can give generic definitions
of data structure assertions (e.g., linked lists, etc.) which do not need to track information
about their own address space. In fact, \emph{only} assertions that explicitly deal with multiple address
spaces need to mention address spaces at all (via the other-space modality).
}

We can prove that this modality follows certain basic laws, showing that its truth is independent of the address
space in which it is considered, that it is distributed over various logical connectives, and that it follows the rule of
consequence.
We call \textsf{vProp} assertions whose truth is independent of the current address space
\textsf{Fact}s; these include other address-space assertions, physical memory points-tos, and register assertions.
\textsf{Fact}s can freely move in and out of other address-space modalities.
\looseness=-1

In general, per-address-space invariants should be collected in a larger
VMM invariant, with individual address spaces' invariants pulled out as needed, such as when proving
soundness of an individual virtual memory access.
However, such larger invariants would contain many kernel-specific properties that are orthogonal
to the fundamental reasoning principles that are the focus of this paper.
We leave such kernel-specific reasoning to future work, but our verification of task switching
(Section \ref{sec:experiment}) demonstrates support for managing multiple address spaces.
\looseness=-1

@@

\subsection{Selected Logical Rules}
\label{sec:selected_rules}
% Per the discussion in Section \ref{sec:issues}, w
As common for assembly-level verification~\add{\cite{Ni2006codeptrs,ni2007contexts}}, we define our logic using Hoare doubles:%\footnote{This
% omits some low-level Iris details (stuckness, observations) that play no meaningful
% role in our development.}
  \[
  \begin{array}{l}
    %\textsf{wpd\_def e s E1 } \Phi \;\mathsf{ rtv } : \textsf{iProp }\Sigma := \\
    \{ \Phi \}_\mathsf{ rtv }\;\textsf{e} : \textsf{iProp }\Sigma := 
   % \qquad
   ((\textsf{cr3} \mapsto_{\textsf{r}} \textsf{rtv} \ast \Phi) \textsf{ rtv}) \wand \textsf{WP e } \{\_, \textsf{True} \}
    \end{array}
  \]
Our Hoare doubles $\{\Phi\}_\textsf{rtv}\;\textsf{e}$ state that the expression (i.e., sequence of instructions)
\textsf{e} are safe to execute (will not fault)
when executed with \textsf{vProp} precondition $\Phi\ast\textsf{cr3}\mapsto_{\textsf{r}} \textsf{rtv}$.
\textsf{WP} is \iris's own weakest precondition modality, unmodified~\cite{jung2018iris}.
Making \textsf{rtv} a parameter to the double (versus a simple registration assertion)
makes it possible to ensure ownership of the \lstinline|cr3| register and its value is accounted for
while avoiding some technical headaches with trying to enforce that $\Phi$ itself contains that.
\looseness=-1
% solves a technical problem with ensuring that the page table root used to evaluate
% the \textsf{vProp} (i.e., evaluating the assertion in the \emph{current}) address space
% is feasible.
% \footnote{Consider the difficulty of selecting the correct page table root value from an arbitrary
% opaque $\Phi$, which may even existentially quantify the page table root. An alternative is to
% require $\Phi$ to have a syntactic form where we can directly extract the value of \lstinline|cr3|,
% but this makes using Iris Proof Mode (IPM)~\cite{Krebbers:2017:IPH:3009837.3009855} with \textsf{vProps}
%   difficult; IPM works for any type matching the signature of an Iris \lstinline|bi|, which includes
%   \textsf{vProp}s, but manually guiding IPM to put an assertion in a specific position over and over adds
%   significant proof burden.
% }

>>>>

We describe a program logic (a separation logic) along the lines suggested earlier, where every assertion is relative
to an address space in which it is interpreted, allowing us to define \emph{virtual points-to} assertions that make claims
about memory locations in a particular address space. Virtual addresses, and even virtual points-to assertions, 
are not tagged with their address spaces in any way. Memory access in this logic is validated through the use
of virtual points-to assertions in preconditions, which guarantee that address translations succeed.
This supports rules for updating not only typical data in memory that happens to be subject to address translation, but \add{also}
manipulation of the page tables themselves via virtual addresses (as demanded by all modern hardware) \replace{also}{and}
via virtual points-to assertions.
To support specifications that deal with multiple address spaces, our logic incorporates a hybrid-style modality
$[r](P)$ to state that an assertion is true in another (assertion-specified) address space rather than the address space
currently active in hardware, which is not only useful for virtual memory manager invariants, but \add{also} critical to reasoning
about change of address space.
By developing this within the \iris framework, we obtain additional features (e.g., fractional permissions) that allow us to verify
some of the most subtle and technically challenging instruction sequences in an OS kernel (Section \ref{sec:experiment}).

% We derive a program logic (a separation logic) supporting the following stances and constraints:
% \begin{enumerate}
% \item \textit{address spaces as modal contexts}: Assertions in our logic are context-dependent,
%   in the sense that their truth depends on which address space they are used in, due to the need to support virtual points-to assertions.
% \item \textit{sharing}: The physical location backing a virtual address's storage is located (during a page table walk) through a 
%       set of physical page-table (L4-L1 page-tables) acceses that are shared amongst different virtual addresses (specifically,
%       those on the same page of memory\footnote{of in the case of L2 or higher levels of tables, within a given broader region.}).
%       This sharing imposes constraints on defining points-to assertions
%       in terms of physical (L4-L1) page-table memory accesses
% \item \textit{context-agnostic-resources}: each virtual address is valid under a certain address space, 
%       but it does not represent this \textit{knowledge} of its address space. 
%       That is, assertions are not explicitly tagged with their address space validity
% \item \textit{updating address-space mappings}: We present logical abstractions to enable 
%       updating not only pages of typical data in memory, but also page tables themselves.\footnote{Prior work relied on unfolding operational semantics
%       to verify page table updates.}
% \item \textit{explicitly-modal assertions}: Our logic includes a means to talk about facts being true
%       in another address space
% \item \textit{address-space switch as changing the "World" of truth}: Switching from one address-space to another logically
%       becomes a simultaneous introduction-and-elimination of a pair of modal assertions (for different address spaces)
% \end{enumerate}

% The idea that the truth of an assertion is relative to an address space has far-reaching consequences.
To support making assertions depend on a choice of address space, we work entirely in a pointwise lifting of \iris's base BI logic,
essentially working with separation logic assertions indexed by a choice of page table root as a $\mathcal{W}_{64}$, which we call $\textsf{vProp }\Sigma$:\footnote{
  \iris experts may notice our \lstinline|-b>| resembles another pointwise lifting already in  \iris~\cite{dang2019rustbelt,dang2022compass}. 
  This similarity is real, but the existing lifting does not appear to work with indexed \coq types like our \lstinline|word n| as a domain.
}
\lstinline[language=Coq]|Definition vProp  $\Sigma$ : bi := word 64 -b> iPropI  $\Sigma$|.
This is the (\rocq) type of assertions in our logic.
Most constructs in \iris's base logic are defined with respect to any BI-algebra (of \coq type \lstinline|bi|), so \add{they} automatically
carry over to our derived logic.
\add{Since \textsf{vProp}s are ``just'' \textsc{Iris} predicates on a choice of page table root,
it may not be obvious what the gain is over simply using \textsc{Iris} directly.
The gain is not expressive power.
Every modal logic can be translated straightforwardly to a logic without modalities
(nearly always a small variation on the \emph{standard translation} from propositional modal logic
to first-order logic~\cite{blackburn2006modal}\footnote{\add{Itself similar to verification condition generation
for weakest precondition calculi and Hoare logics.}}) that can directly address the logic's models ---
but these translations, in addition to making specifications far more verbose,
generally do not preserve the appealing reasoning principles of modal logics.
This becomes \replace{most}{more} apparent when discussing our proof rules.
}

@@

\paragraph{Register points-to} 
The assertion $\textsf{r}\;\mapsto_{\textsf{r}}\{q\}\;\textsf{rv}$ ensures the ownership of the register $\rg$ containing the
value of the register $\rv$.\todo{Ismail, you changed ``the register value $\rv$'' to ``value of the register $\rv$'' and I changed it back. What you wrote treats $\rv$ as the name of the register, not its value.}
The fraction $\qfrac$ with value 1 asserts the unique ownership of the register mapping and grants update permission \replace{on}{to} it;
otherwise, any value $0 < \qfrac <1$ represents partial ownership, granting read-only permission on the mapping.

@@

\lstinline|L4_L1_PointsTo| works by
chaining together the entries for each level, using the sequence of table offsets from the address being translated to index
each table level, and using the physical page address embedded in each entry.\footnote{
  The fractions \lstinline|q1| through \lstinline|q4| represent the fractional ownership of each entry based on how many
  word-aligned addresses might need to share the entry ---  $(\frac{1}{512})^n$ for each level $n$.
}
For example, the first-level address translation to get the L4 entry (\lstinline|l4e|) 
  uses the masks \textsf{l4M52} with the current \lstinline|cr3| to get the physical address of the start of the L4 table
  and \textsf{l4off} with the virtual address being translated to compute the correct offset within that table.
  Then \del{at} that physical location \replace{is}{holds} the appropriate entry in the L4 table \textsf{l4e}.
  Subsequent levels of the page table walk work similarly.
The statement of these assertions is simplified by the use of our split physical points-to assertions, since
each level of tables is page-sized. \footnote{We do not address superpages and hugepages in this paper.}
This helper definition is also more explicit in \textsf{vProp} which binds a value to \lstinline|cr3| and uses it to start the translation process.
\looseness=-1

@@

  We separate the physical page-table walk from the virtual points-to relation, replacing it with a ghost state that merely guarantees that the address translation would succeed.
  \iris includes a \emph{ghost map} construction, which we use to track mappings from virtual addresses to the physical addresses they translate to as a piece of ghost state.
  The map includes, for each key in the map (i.e.,
  each virtual address), a token $\ghostmaptoken{\gamma}{k}{v}$ whose ownership is required to update that key-value pair in the ghost map named $\gamma$. The existence of such a token implies that the actual map $\theta$ tracked by a corresponding $\mathsf{GhostMap}(\gamma,\theta)$
  resource indeed maps $k$ to $v$. These properties are captured by key \iris rules in Figure \ref{fig:ghostmaps}.\footnote{\iris ghost maps lack established notation in the literature,
  but the syntax we use captures the details of \texttt{iris.base\_logic.lib.ghost\_map}.}
  There are other rules, but these two are most important for explaining ghost maps.
  \textsc{GhostMapUpdate} says that ownership of the actual ghost map with ghost name $\gamma$ and map contents $\theta$,
  and a full (non-fractional) token witnessing that $\theta$ maps \textsf{pa} to \textsf{va} permits an update to the ghost map's state,
  changing the map and replacing the token to represent the new value.
  \textsc{GhostMapLookup} allows using the same information to simply conclude that the mapping indicated by the token is true.

@@

  % In Iris this is realized by using an authoritative resource algebra: there is a single \emph{authoritative} global copy of the (ghost)
  % map caching virtual-to-physical address translations, and for each entry a read only \emph{partial} ownership of that key-value pair.
  % The resource algebra itself is instantiated as:
  % \[\mathcal{A}\textsf{bsPTableWalk} \stackrel{\triangle}{=} \textsc{Auth} (\; \mathcal{W}_{64} \;\rightarrow_{\textrm{fin}} \;  ( (\textsc{Frac }, \mathord{+}) \times (\textsc{Agree } \Loc,\mathord{=}) ))\]
For clarity, we refer to the specific ghost map summarizing virtual-to-physical translations by \\
\centerline{$\mathcal{A}\textsf{bsPTableWalk}(\delta,\theta) \stackrel{\triangle}{=} \mathsf{GhostMap}(\delta,\theta)$}\\
(omitting $\delta$ for brevity when only one is in scope)
and keep this in a per-address-space invariant described shortly.
We then replace the physical traversal $\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in Figure \ref{fig:strongvirtualpointsto}'s virtual points-to definition
with ownership of the token \ghostmaptoken{\delta}{\vaddr}{\paddr}, %ghost-map ($ \sumwalkabs\vaddr\qfrac\paddr$),
yielding Figure \ref{fig:virtualpointstosharing}'s definition.
This new definition guarantees that the ghost map contains an mapping from the virtual address ($\vaddr$) to a physical address ($\paddr$),
and thus that the per-address-space invariant will contain the physical resources that guarantee \add{that} the hardware resolves the translation.

  \begin{figure*}
  \footnotesize
\centerline{$
\begin{array}{l}
  \mathcal{I}\textsf{ASpace}(\ptablestore,m)\stackrel{\triangle}{=} \textsf{ASpace\_Lookup}(\ptablestore,m) \ast \\
 \qquad\qquad\qquad \bigast{(\vaddr, \textsf{paddr})\in \ptablestore}{\exists\;(\textsf{l4e l3e l2e, l1e, paddr})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e, paddr})} \\
  \textsf{ where } 
   \textsf{ASpace\_Lookup} (\ptablestore,m) \stackrel{\triangle}{=} \lambda\textsf{ cr3val} \ldotp \; \exists \gammaPred \; \ldotp \ulcorner m \; !!\; \textsf{cr3val} = \textsf{Some } \gammaPred \urcorner \ast
    %\ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}
    \ptableabswalk{\delta,\theta}
\end{array}
$}
\vspace{-1em}
\caption{Per-address-space invariant with a fixed global map of address space names $m$}
  \label{fig:peraspaceinvariant}
\vspace{-1em}
  \end{figure*}

We place the authorative ownership of the ghost translation $\mathcal{A}\textsf{PTableWalk}$ in a per-address-space invariant
$\mathcal{I}$\textsf{ASpace} (Figure \ref{fig:peraspaceinvariant}). 
$\mathcal{I}$\textsf{ASpace} alone allows changes to the page tables
that preserve overall virtual-to-physical translations.
When combined with the
fragment stored in the virtual points-to (Figure \ref{fig:virtualpointstosharing}),
changes in the virtual-to-physical translations become possible without the need to track all virtual points-to assertions.

\todo[inline]{This next paragraph below explains details, but should first explain the big picture: the current address space
is identified by a paddr, so to state the invariant for the address space named by that paddr we need to look up
what invariants should hold, then assert that those invariants do hold.}
In Figure \ref{fig:peraspaceinvariant},
$\theta$ is the logical map from virtual addresses to the physical addresses they should translate to, 
for the \emph{current} address space, corresponding to the  currently installed page tables (the ones
indicated by \lstinline|cr3|). In Figure \ref{fig:virtualpointstosharing}, the existentially quantified $\delta$
is the ghost name for this map (again, for the current address space).
Because our logic is the first to address the existence of multiple address spaces,
it must ensure that $\theta$ actually corresponds to the intended ghost map for the current address space.
This requires looking up the ghost name (i.e., $\delta$) for the current address space,
which is handled by $m$ --- another logical map (tied to ghost state shortly)
from various page table roots to ghost names for the per-address-space ghost maps discussed above.
$\mathcal{I}\textsf{ASpace}(\theta,m)$ then stores the physical resources for the current address space mappings,
ensuring that the current address space is valid (has an entry in $m$ whose corresponding
authoritative map matches $\theta$), and asserting ownership of the appropriate fractional ownership
of the physical page table walk for each virtual-to-physical translation in $\Theta$ (for the current address space).

@@

\subsection{Address Space Management}
\label{sec:aspacemanagement}
% So far, we have introduced logical abstractions for a single address space, but VMMs
Real VMMs must
 handle more than one address space.
Doing so requires a way to talk about other address spaces, and means to switch address spaces.
\begin{figure}
\footnotesize
\begin{mathpar}
    \mbox{$[r](\mathsf{P}) \;: \mathsf{vProp} \; \Sigma \; \stackrel{\triangle}{=} \; \lambda \_,\; \mathsf{P}~\mathsf{r} \vspace{0.5em}$}
    \and
    \mbox{$\textsf{Fact P} \stackrel{\triangle}{=} \;\forall \; ,r \; r' \ldotp  \textsf{P r} \dashv\vdash \textsf{P r'}$}
    \and
    \inferrule*{ }{\textsf{Fact}\;[r](\textsf{P}) }
    \and
    \inferrule*{ }{\textsf{Fact}\;(\textsf{r}\;\mapsto_{\textsf{r}}\{q\}\;\textsf{rv})}
    \and
    \inferrule{ }{\textsf{Fact}\;(\textsf{w} \mapsto_{\textsf{p}} \{q\} \textsf{ v} )}
    \and
    \inferrule{ }{(\textsf{P} \vdash \textsf{Q}) \vdash  ([r](\textsf{P}) \vdash  [r](\textsf{Q}))}
    \and
    \inferrule{ }{[r](\textsf{P} \ast \textsf{Q}) \dashv\vdash ([r](\textsf{P}) \ast [r](\textsf{Q})}\vspace{0.5em}
    \and
    \inferrule{ }{[r](\textsf{P}\land \textsf{Q}) \dashv\vdash [r](\textsf{P})\land [r](\textsf{Q})}
    \and
    \inferrule{ }{[r](\textsf{P}\lor \textsf{Q}) \dashv\vdash [r](\textsf{P})\lor [r](\textsf{Q})}
    \and
    \inferrule{ }{\textsf{Fact P} \vdash  [r](\textsf{P}) \dashv\vdash (\textsf{P})} %(\textsf{vProp }\Sigma))\\
\end{mathpar}
\vspace{-1em}
  \caption{Other-space Modality and Its Laws}
  \label{fig:modaldef}
\vspace{-1em}
  \end{figure}
Figure \ref{fig:modaldef} gives the definition of our modal operator for asserting the truth of a modal
(address-space-contingent) assertion \emph{in another address space}, which we call
the \emph{other-space} modality. The definition itself is not
particularly surprising --- as our modal assertions are semantically predicates on a page table root (physical)
address, the assertion $[r](P)$ is a modal assertion that ignores the (implicit) current page table root,
and evaluates the truth of $P$ as if $r$ were the page table root. 
The novelty here is not in the details of the definition, but in recognizing that this is the right way to deal with
multiple address spaces, and working out how to support interaction of multiple address spaces (discussed in the next section).
\add{The modal assertions, together with the other-space modality, mean we can give generic definitions
of data structure assertions (e.g., linked lists, etc.) which do not need to track information
about their own address space. In fact, \emph{only} assertions that explicitly deal with multiple address
spaces need to mention address spaces at all (via the other-space modality).
}

We can prove that this modality follows certain basic laws, showing that its truth is independent of the address
space in which it is considered, \add{that} it distributes over various logical connectives, and \add{that} it follows the rule of
consequence.
We call \textsf{vProp} assertions whose truth is independent of the current address space
\textsf{Fact}s; these include other-space assertions, physical memory points-tos, and register assertions.
\textsf{Fact}s can \add{freely} move in and out of other address space modalities\del{ freely}.
\looseness=-1

In general, per-address-space invariants should be collected in a larger
VMM invariant, with individual address spaces' invariants pulled out as needed, such as when proving
soundness of an individual virtual memory access.
\replace{But}{However,} such larger invariants would contain many kernel-specific properties that are orthogonal
to the fundamental reasoning principles that are the focus of this paper.
We leave such kernel-specific reasoning to future work, but our verification of task switching
(Section \ref{sec:experiment}) demonstrates support for managing multiple address spaces.
\looseness=-1

@@

\subsection{Selected Logical Rules}
\label{sec:selected_rules}
% Per the discussion in Section \ref{sec:issues}, w
As common for assembly-level verification~\add{\cite{Ni2006codeptrs,ni2007contexts}}, we define our logic using Hoare doubles:%\footnote{This
% omits some low-level Iris details (stuckness, observations) that play no meaningful
% role in our development.}
  \[
  \begin{array}{l}
    %\textsf{wpd\_def e s E1 } \Phi \;\mathsf{ rtv } : \textsf{iProp }\Sigma := \\
    \{ \Phi \}_\mathsf{ rtv }\;\textsf{e} : \textsf{iProp }\Sigma := 
   % \qquad
   ((\textsf{cr3} \mapsto_{\textsf{r}} \textsf{rtv} \ast \Phi) \textsf{ rtv}) \wand \textsf{WP e } \{\_, \textsf{True} \}
    \end{array}
  \]
Our Hoare doubles $\{\Phi\}_\textsf{rtv}\;\textsf{e}$ state that the expression (i.e., sequence of instructions)
\textsf{e} are safe to execute (will not fault)
when executed with \textsf{vProp} precondition $\Phi\ast\textsf{cr3}\mapsto_{\textsf{r}} \textsf{rtv}$.
\textsf{WP} is \iris's own weakest precondition modality, unmodified~\cite{jung2018iris}.
Making \textsf{rtv} a parameter to the double (vs.\ a simple register assertion)
makes it possible to ensure ownership of the \lstinline|cr3| register and its value is accounted for
while avoiding some technical headaches with trying to enforce that $\Phi$ itself contains that.
\looseness=-1
% solves a technical problem with ensuring that the page table root used to evaluate
% the \textsf{vProp} (i.e., evaluating the assertion in the \emph{current}) address space
% is feasible.
% \footnote{Consider the difficulty of selecting the correct page table root value from an arbitrary
% opaque $\Phi$, which may even existentially quantify the page table root. An alternative is to
% require $\Phi$ to have a syntactic form where we can directly extract the value of \lstinline|cr3|,
% but this makes using Iris Proof Mode (IPM)~\cite{Krebbers:2017:IPH:3009837.3009855} with \textsf{vProps}
%   difficult; IPM works for any type matching the signature of an Iris \lstinline|bi|, which includes
%   \textsf{vProp}s, but manually guiding IPM to put an assertion in a specific position over and over adds
%   significant proof burden.
% }
